Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-10 write-ups/PLDI/paper.tex
Progress

\paragraph{Id functions}

@@

In general, an OpenCL kernel is organised as a multi-dimensional grid
of multi-dimensional workgroups, and the above id functions each take
an integer dimension argument.  For simplicity, we consider only the
one-dimensional case throughout the paper, which captures all the
GPU-accelerated irregular algorithms we are aware of, and use
e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.

@@

We now explain the semantics of these language extensions, after which
we explain how the work stealing and graph traveral algorithms of
Section~\ref{sec:background} can be turned into cooperative kernels
that use the extensions.

@@

A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new $\transmit$ qualifier.  By
annotating a variable with $\transmit$, the programmer indicates that
when a workgroup spawns new workgroups by calling $\offerfork$, the
workgroup should transmit the current value of this variable to serve
as an initial value for the new workgroups.  We detail the semantics
for this when we describe $\offerfork$ below.

\ADComment{Remove?}
Only private memory variables can be annotated with $\transmit$,
which may include pointers to local or global memory so long as the
pointer variables themselves are stored in private memory.  In
particular, variables and arrays in local memory cannot be annotated
with $\transmit$.  \ADComment{The preceding is what we have
  implemented, but actually it is pretty clear how we could handle
  local memory restoration if we wanted to; it's just that none of our
  examples needed it.}

@@

If the host application launches a cooperative kernel specifying that
$N$ workgroups should execute the kernel, then when the kernel
commences execution it should be executed by some number of workgroups
$M$, with $1\leq M \leq N$.  The requested number of workgroups is
called the \emph{workgroup limit} for the cooperative kernel, and $M$
is the number of \emph{active workgroups}.  As explained below, the
number $M$ of active workgroups may change during the lifetime of a
cooperative kernel.

@@

The initial workgroups have consecutive workgroup ids in the range
$[0, M-1]$.

The OpenCL primitive $\getgroupid{0}$ retrieves the id of a
workgroup as usual.  When executed by a cooperative kernel, the OpenCL
primitive $\getnumgroups{0}$ returns $M$, the current number
of active workgroups.  Because the ND-ranges of cooperative kernels
are one-dimensional, $\getgroupid{\mathit{i}}$ and
$\getnumgroups{\mathit{i}}$ return 0 and 1, respectively, for $i >
0$.

\ADComment{What happens to $M$ when a workgroup reaches the end of a
  cooperative kernel?  Feels like $M$ should decrease, but what does
  this mean for barrier-only kernels?  Do they need to account for the
  number of active workgroups changing during the final barrier
  interval?  An alternative is to say that there is an implicit global barrier at kernel exit.}

@@

\ADComment{Don't allow zero to be killed.}

\ADComment{Explain that the ``last one dies'' semantics is convenient because it maintains contiguous ids, but other options are possible and may be more efficient sometimes.}

Similar to the OpemncL workgroup $\mathsf{barrier}$ primitive, our first additional language primitive, $\offerkill$, is a
workgroup-level function (see Section~\ref{sec:background}).

Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$,
i.e.\ the workgroup's id is not the largest among the active
workgroups, then $\offerkill$ is a no-op.  If instead $m = M-1$, there
is a non-deterministic choice between (a) $\offerkill$ executing as a
no-op, or (b) execution of the workgroup ceasing and the number of
active workgroups $M$ being decremented by one.  Case (b) corresponds
to the scheduler accepting the workgroup's offer to cease execution,
allowing the scheduler to schedule some other task on the compute unit
on which the workgroup was executing.

The semantics of $\offerkill$ is atomic.  That is, if multiple
workgroups reach $\offerkill$ simultaneously, the semantics is as if
each workgroup executes $\offerkill$ in some arbitrary sequential
order.  \ADComment{Should we say something about linearisability?}

@@

Our second new language primitive, $\offerfork$, is also a workgroup-level
function.

@@

The $k$ new workgroups commence execution at the program point
immediately following the $\offerfork$ call.  The private variables of
each thread in a new workgroup are all uninitialised, as are any local
variables and arrays shared across the workgroup; reading from these
locations without initialising them is an undefined behaviour.  There
are two exceptions to this:

@@

\item because the parameters to a cooperative kernel are immutable,
  the new threads have access to these parameters and can safely read
  from them;

@@

@@

\ADComment{Remark on our decision to have every thread inherit the
  state of the master thread from workgroup $m$.}

@@

@@

@@

\section{Semantics}

>>>>

\paragraph{Identity and size functions}

@@

In general, an OpenCL kernel is organised as a multi-dimensional grid
of multi-dimensional workgroups, and the above functions each take
an integer dimension argument.  For simplicity, we consider only the
one-dimensional case throughout the paper, which captures all the
GPU-accelerated irregular algorithms we are aware of, and use
e.g.\ $\getglobalid()$ to mean $\getglobalid(0)$.

@@

We now explain the semantics of these language extensions, after which
we explain how the work stealing and graph traveral algorithms of
Section~\ref{sec:background} can be turned into cooperative kernels
that use the extensions.  We present the semantics intuitively, using English.
In Appendix~\ref{appendix:semantics} we present an abstract operational semantics for our language extensions in a simple GPU-like programming model.

@@

A variable declared in the root scope of the cooperative kernel can
optionally be annotated with a new $\transmit$ qualifier.  By
annotating a variable with $\transmit$, the programmer indicates that
when a workgroup spawns new workgroups by calling $\offerfork$, the
workgroup should transmit its current value for the variable to serve
as an initial value for the variable for the threads of the new workgroups.  We detail the semantics
for this when we describe $\offerfork$ below.

@@

If the host application launches a cooperative kernel requesting $N$
workgroups, this indicates that the kernel should be executed with a
maximum of $N$ workgroups, and that as many workgroups as possible, up
to this limit, are desired.  However, the scheduler may initially
schedule fewer than $N$ workgroups, and as explained below the number
of workgroups that execute the cooperative kernel can change during
the lifetime of the kernel.

The workgroups executing the kernel are called \emph{active
  workgroups}, and $M$ denotes the number of active workgroups.
Initially, the number of active workgroups $M$ is initialised to some
integer between $1$ and $N$.  The initial active workgroups have
consecutive workgroup ids in the range $[0, M-1]$.

The $\getgroupid()$ function retrieves the id of a workgroup as usual
(see Section~\ref{sec:background}).  When executed by a cooperative
kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups()$ for
regular kernels, which returns the fixed number of workgroups that
were requested at kernel launch time.

@@

@@

The $\offerkill$ primitive allows the cooperative kernel to return
compute units to the scheduler by offering to sacrifice workgroups.
The idea is that while having the scheduler be permitted to terminate
the execution of workgroups in an arbitary fashion would be drastic,
the kernel may contain identifiable points of execution at which a
workgroup could gracefully leave the computation.  We give examples of
these for work stealing and graph traversal below.

Similar to the OpenCL workgroup $\mathsf{barrier}$ primitive,
$\offerkill$, is a workgroup-level function---it must be encountered
uniformly by all threads in a workgroup (see
Section~\ref{sec:background}).

Suppose a workgroup with id $m$ executes $\offerkill$.  If $m < M-1$
or $M=1$, i.e.\ the workgroup's id is not the largest among the active
workgroups, or if only a single workgroup is executing the cooperative
kernel, then $\offerkill$ is a no-op.  If instead $M > 1$ and $m =
M-1$, there is a non-deterministic choice between (a) $\offerkill$
executing as a no-op, or (b) execution of the workgroup ceasing and
the number of active workgroups $M$ being decremented by one.  Case
(b) corresponds to the scheduler accepting the workgroup's offer to
cease execution, allowing the scheduler to schedule some other task on
the compute unit on which the workgroup was executing.

The semantics of $\offerkill$ is atomic.  That is, if multiple
workgroups reach $\offerkill$ simultaneously, the semantics is as if
each workgroup executes $\offerkill$ in some arbitrary sequential
order.

In this formulation of cooperative kernels, only the active workgroup
with the largest id can be killed, and workgroup 0 can never be
killed.  We discuss the rationale for these choices, as well as
possible alternative choices, in
Section~\ref{sec:semanticalternatives}.

@@

The $\offerfork$ primitive allows the cooperative kernel to indicate
to the scheduler that it could benefit from, and is prepared to
handle, additional workgroups joining the computation.  Recall that
the cooperative kernel was launched with a limit of $N$ workgroups,
indicating the number of workgroups that would ideally execute the
kernel.  However, the number of active workgroups, $M$, may be smaller
than $N$, either because (due to competing workloads) the scheduler
did not provide $N$ workgroups initially, or because the kernel has
given up some number of workgroups via $\offerkill$ calls.  Through
$\offerfork$, the kernel and scheduler can work together to allow new
workgroups to join the computation at an appropriate point and with
appropriate state.

Like $\offerkill$, $\offerfork$ is a workgroup-level function.

@@

The $k$ new workgroups commence execution at the program point
immediately following the $\offerfork$ call.  The variables that
describe the state of a thread are all uninitialised for the threads
in the new workgroups; reading from these variables without first
initialising them is an undefined behaviour.  There are two exceptions
to this:

@@

\item because the parameters to a cooperative kernel are immutable,
  the new threads have access to these parameters as part of their
  local state and can safely read from them;

@@

In effect, thread 0 of the spawning workgroup transmits the relevant
portion of its local state to the new threads.  We discuss in
Section~\ref{sec:semanticalternatives} the rationale for having all
new threads obtain this local state portion from thread 0, as well as
our reasons for selecting which variables to transmit via annotations
rather than transmitting the entire thread state.

@@

@@

\ADComment{Allowing arbitrary workgroups to be killed (not just the largest workgroup)}

\ADComment{Having thread 0 transmit its state, and transmitting only annotated variables.}

@@

\appendix

\section{Operational Semantics for Cooperative Kernels}\label{appendix:semantics}

\ADComment{TODO: write it up properly.}

\ADComment{TODO: remark that we do not model transmit faithfully.}

@@
